\bf A Comparison of Parametric and Permutation Tests for Regression Analysis of Randomized Experiments

### 2017-10-02 ms/ms.tex
More edits to MS addressing reviewer comments

\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{url} % not crucial - just used below for the URL 
\usepackage{color}
 \usepackage{multirow}

@@

\bigskip
\begin{abstract}
% 200 words or less
Hypothesis tests based on linear models are widely accepted by organizations that regulate clinical trials.
These tests are derived using strong assumptions about the data-generating process so that the resulting inference can be based on parametric distributions.
Because these methods are well understood and robust, they are sometimes applied to data that depart from assumptions, such as ordinal integer scores.
Permutation tests are a nonparametric alternative that require minimal assumptions which are often guaranteed by the randomization that was conducted.
We compare analysis of covariance (ANCOVA), a special case of linear regression that incorporates stratification, to several permutation tests based on linear models that control for pretreatment covariates.
In simulations using a variety of data-generating processes, some of which violate the parametric assumptions,
the permutation tests maintain power comparable to ANCOVA.
We illustrate the use of these permutation tests alongside ANCOVA with data from a clinical trial comparing the effectiveness of two treatments for gastroesophageal reflux disease.
Given the considerable costs and scientific importance of clinical trials, one may want to include an additional nonparametric method, such as a linear model permutation test, as a robustness check on the statistical inference for the main study endpoints.
\end{abstract}

@@

\newpage
\spacingset{1.45} % DON'T change the spacing!
\section{Background}
A hypothesis test is a statistical method for determining whether observed data is consistent with a belief about the process that generated the data.
Medical experiments use hypothesis testing to assess the evidence that a treatment affects one or more clinically relevant outcomes.
The simplest version of this experiment involves randomly assigning two treatments to a fixed number of individuals in a group and measuring a single outcome, and has been studied for nearly a century (see \cite{fisher_design_1935} and \cite[1990 translation]{neyman_application_1923} for early references).
One can conduct hypothesis tests and construct confidence intervals for the estimated treatment effect
by exploiting the fact that the difference in average outcomes between the two treatment groups is asymptotically normal with a variance that can be estimated from the data.

Randomized experiments in the real world are rarely this simple:
if individuals are heterogeneous before the study, then their outcomes may differ for reasons besides the treatment;
there may be more than two treatments under study;
patients may not take the treatment they are assigned or may drop out of the study;
and peer effects, the influence of other members of a treatment group, may affect outcomes beyond the treatment itself.
Techniques have been developed to deal with each of these issues, invoking additional assumptions to make valid inference possible.
In this paper, we focus on the issue of heterogeneous pretreatment characteristics and assume that the other issues are not present.

Random assignment of treatment ensures that pretreatment covariates are balanced between treatment groups on average, across all possible randomizations.
However, in any particular randomization, there may be imbalances.
If the imbalanced variables are associated with the outcome, then even when treatment has no effect, there may be differences in outcomes between treatment groups.
Adjusting for such covariates can reduce the variability of treatment effect estimates and yield more powerful hypothesis tests.

Stratification, sometimes called blocking, is one method to control for covariates that are known a priori to be associated with the outcome.
Strata are groups of individuals with similar levels of a covariate.
These groups are defined during the design stage (i.e. before outcome data are collected).
Random assignment of treatments is conducted within each stratum, independently across strata.
This guarantees that the stratification variable is balanced between treatment groups.
A common stratification variable in clinical experiments is location:
individuals often come from many sites because it is difficult to recruit a sufficient number of participants at one doctor's office or hospital,
especially when the object of study is a rare disease or a rare outcome.
Stratification can have the same beneficial effects as controlling for other pretreatment covariates, 
as outcomes may vary across strata due to factors unrelated to individuals' treatment and covariates.

Linear regression is another way to control for imbalanced baseline covariates.
It is done during the data analysis stage.
In its simplest form, linear regression projects the outcomes onto the plane that best summarizes each variable's relationship to the outcome.
The coefficient of any particular covariate answers the question, if we were to hold fixed all other variables and increase this variable by one unit, how much would we expect the outcome to change?
This model posits a linear relationship between covariates, treatment, and outcome; 
if the true relationship is not linear, then linear regression gives the best linear approximation to the conditional expectation of the outcome.
It is standard to use analysis of covariance (ANCOVA) to account for stratification in a linear model.
ANCOVA is a particular case of linear regression that allows the mean outcome to vary from stratum to stratum.
This amounts to fitting a plane to each stratum, with the constraint that they have a common slope.

Hypothesis testing of estimated coefficients requires even stronger assumptions than linearity, namely Gaussian, homoskedastic errors.
When this holds, a regression coefficient scaled by its estimated variance follows Student's $t$ distribution.
Thus, if we believe the linear model is correctly specified, a hypothesis test for a treatment effect amounts to a hypothesis test of the coefficient for treatment in the ANCOVA model, and can be evaluated analytically and efficiently.
However, the modeling assumptions are clearly violated in many cases:
when the data are discrete or ordinal, 
when the treatment has a differential effect across subgroups of individuals and no interaction terms are included, 
and when the variance of outcomes differs across strata.
Furthermore, the hypothesis test implicitly assumes that the data were sampled at random from some underlying population, when in fact,
medical experimenters rarely recruit patients this way (\cite{ludbrook_why_1998}).

@@

\subsection{Notation}
Suppose we have a finite population of $N$ individuals.
Individuals are grouped into strata indexed by $j = 1, \dots, J$, with $n_j$ individuals in stratum $j$ and $\sum_{j=1}^J n_j = N$.
Of the $n_j$ subjects in stratum $j$, $n_{tj}$ are assigned treatment 1, while the remaining $n_j - n_{tj}$ are assigned treatment 0.
Let $Z_{ij}$ indicate the treatment assigned to individual $i$ in stratum $j$.

@@

Furthermore, we observe a covariate $X_{ij}$ that may be associated with the outcome.
For expository clarity, we suppose that $X$ is univariate, but all results are easily extended to the case when $X$ is multivariate.
Suppose further that there is a categorical pretreatment variable with $J$ levels used to stratify individuals.

We are interested in the effect of treatment, measured as differences in potential outcomes $Y_{ij}(1) - Y_{ij}(0)$.
We may never learn this difference for any particular individual.
However, a tractable problem is to estimate the mean difference in some group, such as the study sample or in a target population.
The chosen method of analysis determines which function of differences is considered.
Various functions of potential outcomes may be of clinical interest; which one prefers depends on the goal of the study.

We study hypothesis testing for whether these differences are nonzero using parametric ANCOVA and its permutation counterparts, assuming this potential outcomes framework throughout.
Other valid methods for comparing two groups include using a two-sample $t$-test to test the difference in two means from normal distributions,
the Wilcoxon rank sum test to test for differences in the medians of two independent groups, 
and the Kolmogorov-Smirnov test and receiver operating characteristic curve analyses to test whether two groups have different distribution functions (\cite{lehmann_nonparametrics_1975,  vexler_statistical_2016}).
We focus on testing using the linear model as this is standard in clinical trials, requires fewer distributional assumptions on the data when using the potential outcomes framework, and incorporates control variables to increase power.

@@

To carry out the standard parametric hypothesis test for a linear model, the following assumptions are needed (\cite{freedman_statistical_2005}):
\begin{enumerate}
\item \textbf{Linearity:} The data $Y$ are related to $X$ and $Z$ linearly.
\item \textbf{Constant slopes:} Stratum membership only affects the intercept $\alpha_j$, not the slopes $\beta$ and $\gamma$.
\item \textbf{IID Errors:} The $\varepsilon_{ij}$ are independent and identically distributed with mean $0$ and common variance $\sigma^2$.
\item \textbf{Independence:} If $X$ is random, $\varepsilon$ is statistically independent of $X$.
\item \textbf{Normality:} The errors are normally distributed.
\end{enumerate}

The coefficients are estimated using least squares (or equivalently, by maximizing the likelihood).
The estimated coefficient $\hat{\gamma}$ is the estimated average treatment effect. 
This procedure also yields an estimate $\hat{\sigma}_{\hat{\gamma}}^2$ of the variance of $\hat{\gamma}$.
Under the null hypothesis, the test statistic 
$$ T = \frac{\hat{\gamma}}{\sqrt{ \hat{\sigma}_{\hat{\gamma}}^2}}$$
follows the Student $t$ distribution with degrees of freedom equal to the number of observations minus the number of parameters estimated (in this case, $N - J - 2$).
The $p$-value for this hypothesis test is the probability, assuming the null hypothesis of zero coefficient is true, that a value drawn from the $t$ distribution is larger in magnitude than the observed $T$.
This test is equivalent to the $F$-test.
When the model assumptions are true, it is uniformly most powerful.

However, in the context of a randomized experiment, these assumptions will not be satisfied.
The estimated treatment effect $\hat{\gamma}$ and nominal standard errors $\hat{\sigma}_{\hat{\gamma}}$ can be severely biased (\cite{freedman_regression_2008, lin_agnostic_2013}).
\cite{lin_agnostic_2013} shows that running a regression with a full set of treatment and covariate interaction terms cannot hurt asymptotic precision, and using the Huber-White sandwich standard errors can yield asymptotically valid confidence intervals.
In small samples, the bias may still be substantial.
\cite{miratrix_adjusting_2013} show that post-stratification, estimating treatment effects within groups of similar individuals defined after data are collected, can ameliorate this bias.
This is equivalent to estimating a fully saturated linear model with interaction terms for treatment and stratum membership.
The linear ANCOVA model does not account for variation in the treatment effect across strata in this way;
if the difference in potential outcomes is heterogeneous across strata, then the coefficient $\gamma$ may be attenuated towards 0.

\subsection{Stratified permutation test}\label{subsec:strat_perm_test}
In randomization inference, the only random quantities are those involving $Z$, the vector of treatment assignments.
The exact inference is derived from the randomization distribution of statistics involving $Z$.
In particular, the potential outcomes and covariates $(Y_{ij}(1), Y_{ij}(0), X_{ij})$ are fixed in the finite population under study, while the observed response $Y_{ij}$ changes with $Z_{ij}$.

@@

In a randomized experiment with complete randomization, any randomization that assigns $n_j$ individuals to treatment 1 in stratum $j$ for each stratum has equal probability.
We can construct the permutation distribution of any statistic under the null hypothesis by imputing the unobserved potential outcomes using the sharp null hypothesis and by re-randomizing treatment assignments using this principle of equal probabilities.
The experimental design specifies the permutation scheme; no additional assumptions are necessary.

@@

To avoid this, one may want to account for the stratification. 
One way to construct a stratified test is to use the same stratified permutation scheme but a different statistic which combines the stratum-specific statistics into a single value, for instance taking the sum of their absolute values.
Taking the absolute value before summing ensures that effects with different signs do not cancel each other out.
Another way to construct a test is to use the nonparametric combination (NPC) framework proposed by \citet{pesarin_permutation_2010}.
In this framework, the ``global'' null hypothesis of no effect whatsoever is decomposed into the intersection of ``partial'' null hypotheses of no treatment effect within each stratum.
For NPC, one first conducts each partial test separately, then combines their $p$-values in a way that preserves dependencies.
This method is equivalent to combining stratum statistics directly when data are from a randomized experiment, because treatment is independent across strata.

\subsection{Permutation tests with the linear model}\label{subsec:lm_perm_tests}
The tests in Section~\ref{subsec:strat_perm_test} only use information on the treatment and the outcome.
However, experimenters record additional covariates that may be predictive of the outcome.
A test that incorporates these covariates may reduce the variance of the statistic and have more power.
The permutation tests in this section use the linear model in Equation~\ref{eqn:ancova}, but make different assumptions about the data.
In a randomized experiment, the treatment assignment is independent of covariates, errors, and potential outcomes,
making several variables exchangeable.
We show two permutation tests developed in this framework.

The first linear model permutation test is a variation on the stratified permutation test described in Section~\ref{subsec:strat_perm_test}.
The only difference is the test statistic:
this test uses the $t$ statistic for the regression coefficient of $Z$ in the linear model in Equation~\ref{eqn:ancova}.
Rather than using the $t$ distribution to calculate the significance of the observed $t$ statistic, we construct the permutation distribution of the $t$ statistic.
Again, the experimental design specifies the randomization distribution and no additional assumptions are needed:
$Z$ is independent of $(Y(1), Y(0), X)$ conditional on strata.

@@

If the null hypothesis is true, then $\gamma = 0$ and $\varepsilon_{ij} = Y_{ij} - \alpha_{ij} - \beta X_{ij}$ for all $i = 1, \dots, n_j, j = 1, \dots, J$.
Therefore, we may estimate the errors by $\hat{\varepsilon} =Y - \hat{Y}$, where $\hat{Y}$ is the vector of fitted values from the regression of $Y$ on $X$ but not $Z$.
The $\hat{\varepsilon}$ approximate the true errors $\varepsilon$ from the true data-generating process, assuming that the null hypothesis is true.
Furthermore, under the null hypothesis, the $\varepsilon$ are independent of $Z$ within strata. 
Thus, we can carry out an approximate permutation test by permuting the estimated $\hat{\varepsilon}$ within strata, independently across strata.

The test has two important assumptions:
the relationship between $Y$ and $X$ is linear and errors are IID within strata.
Together, these conditions ensure that $\varepsilon$ are exchangeable against $X$ within strata.
In practice, we must estimate $\alpha$; using $\hat{\alpha}$ to estimate $\hat{Y}$ yields an approximate permutation test.
The $\hat{\varepsilon}$ may not even be approximately exchangeable within strata if the regression does not describe the relation between $Y$ and $X$ well.
For this reason, \cite{freedman_nonstochastic_1983} advise against using the method if there are large outlier values in the covariate $X$ or when $X$ and $Z$ are highly colinear.
Randomization does not guarantee this additional exchangeability assumption, so it should be checked.
One way, suggested by \cite{freedman_nonstochastic_1983}, is to visually inspect a scatterplot of the residuals against $X$.
One may also run a hypothesis test for association, such as a stratified permutation test using the correlation as a statistic, independently within each stratum and use the nonparametric combination of tests to obtain a single $p$-value (\cite{pesarin_permutation_2010}).

@@

We simulated data from a randomized experiment using several different data-generating processes.
We applied the tests to the data and compared their empirical power over repeated random treatment assignments and random errors.
We compared the following five tests:
the $t$ test from the parametric ANCOVA,
a stratified permutation test using the difference in mean outcomes
 (called ``stratified permutation'' in what follows),
a stratified permutation test using the mean change scores, defined as the difference between the baseline measure and the outcome (called ``differenced permutation'' in what follows),
a stratified permutation test based on the $t$ statistic from the linear regression of outcome on control variables (called ``LM permutation'' in what follows),
and the Freedman-Lane permutation test.
The simulations are included in three supplementary files.

@@

In the first set of simulations, we generated continuous potential outcomes.
We used two scenarios.
In the first, the treatment effect was constant across strata.
We drew a latent random variate $v_{ij}$ from the uniform distribution on $[-4, 4]$.
In the second scenario, the treatment effect varied across strata, and we drew the latent random variable according to
\begin{displaymath}
   v_{ij} \sim \left\{
     \begin{array}{lr}
       \text{Unif}[-4, -1] & : j=1\\
       \text{Unif}[-1, 1] & : j=2\\
       \text{Unif}[1, 4] & : j=3\\
     \end{array}
   \right.
\end{displaymath}

@@

The baseline value for individual $i,j$ was 
$$ X_{ij} = \frac{\gamma e^{v_{ij}} + e^{v_{ij}/2}}{2} + \varepsilon_{ij}$$.

Then we generated potential outcomes as

@@

We included $n_j=16$ individuals per stratum and treatment assignment was balanced, i.e. 8 people received each treatment in each stratum.
After sampling $(v_{ij}, \varepsilon_{ij}, \delta_{ij})$, we regenerated $Z$ and the potential outcomes $Y(Z)$ 10,000 times.
We repeated this procedure for each distribution of latent variables $v$ and of the errors $\varepsilon$ and $\delta$.

The treatment effect for individual $(i, j)$ was $\gamma e^{v_{ij}}$.
We tested whether the treatment had a nonzero effect in the sample using each of the tests described.

\begin{figure}[h]
\centering
\includegraphics[width = \textwidth]{fig/continuous_outcomes_simulation_power.pdf}
\caption{Empirical power curves for the continuous simulated data. The left panel shows power when the treatment effect was constant across strata and the right panel shows power when the treatment effect was heterogeneous.}
\label{fig:continuous_outcomes_sim_power}
\end{figure}

>>>>

\usepackage{amsmath, amsthm}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{url} % not crucial - just used below for the URL 
\usepackage{color}
 \usepackage{multirow}

@@

\bigskip
\begin{abstract}
% 200 words or less
Hypothesis tests based on linear models are widely accepted by organizations that regulate clinical trials.
These tests are derived using strong assumptions about the data-generating process so that the resulting inference can be based on parametric distributions.
Because these methods are well understood and robust, they are sometimes applied to data that depart from assumptions, such as ordinal integer scores.
Permutation tests are a nonparametric alternative that require minimal assumptions which are often guaranteed by the randomization that was conducted.
We compare analysis of covariance (ANCOVA), a special case of linear regression that incorporates stratification, to several permutation tests based on linear models that control for pretreatment covariates.
In several simulations of randomized experiments using a model in which certain regression assumptions are violated,
the permutation tests maintain power comparable to ANCOVA.
We illustrate the use of these permutation tests alongside ANCOVA using data from a clinical trial comparing the effectiveness of two treatments for gastroesophageal reflux disease.
Given the considerable costs and scientific importance of clinical trials, an additional nonparametric method, such as a linear model permutation test, may serve as a robustness check on the statistical inference for the main study endpoints.
\end{abstract}

@@

\newpage
\spacingset{1.45} % DON'T change the spacing!
\section{Background}
A hypothesis test is a statistical method for determining whether observed data is consistent with a belief about the process that generated the data.
Medical experiments use hypothesis testing to assess the evidence that a treatment affects one or more clinically relevant outcomes.
The simplest version of this experiment has been studied for nearly a century (see \cite{fisher_design_1935} and \cite[1990 translation]{neyman_application_1923} for early references).
This experiment involves randomly assigning two treatments to a fixed number of individuals in a group and measuring a single outcome.
One can conduct hypothesis tests and construct confidence intervals for the estimated treatment effect
by exploiting the fact that the difference in average outcomes between the two treatment groups is asymptotically normal with a variance that can be estimated from the data.

%Randomized experiments in the real world are rarely this simple:
%if individuals are heterogeneous before the study, then their outcomes may differ for reasons besides the treatment;
%there may be more than two treatments under study;
%patients may not take the treatment they are assigned or may drop out of the study;
%and peer effects, the influence of other members of a treatment group, may affect outcomes beyond the treatment itself.
%Techniques have been developed to deal with each of these issues, invoking additional assumptions to make valid inference possible.
%In this paper, we focus on the issue of heterogeneous pretreatment characteristics and assume that the other issues are not present.

The goal of a randomized experiment is to draw causal inferences about the efficacy of a treatment.
Thus, it makes sense to analyze experiments in the context of potential outcomes, where every individual has a counterfactual outcome for each of the treatments (\cite{holland_1986_statistics})
Experiments are an attempt to estimate the counterfactual outcomes by randomly assigning treatments.
Random assignment of treatment ensures that pretreatment covariates are balanced between treatment groups on average, across all possible randomizations, making treatment groups comparable to each other.

However, in any particular randomization, there may be imbalances.
If the imbalanced variables are associated with the outcome, then even when treatment has no effect, there may be differences in outcomes between treatment groups.
Adjusting for such covariates can reduce the variability of treatment effect estimates and yield more powerful hypothesis tests.

Stratification, sometimes called blocking, is one method to control for covariates that are known a priori to be associated with the outcome.
Strata are groups of individuals with similar levels of a covariate.
These groups are defined during the design stage (i.e. before outcome data are collected).
Random assignment of treatments is conducted within each stratum, independently across strata.
This guarantees that the stratification variable is balanced between treatment groups.
A common stratification variable in clinical experiments is location:
individuals often come from many locations because it is difficult to recruit a sufficient number of participants at one site,
especially when the object of study is a rare disease or a rare outcome.

Linear regression is another method to control for imbalanced baseline covariates.
It is done during the data analysis stage.
In its simplest form, linear regression projects the outcomes onto the plane that best summarizes each variable's relationship to the outcome.
The coefficient of any particular covariate answers the question, if we were to hold fixed all other variables and increase this variable by one unit, how much would we expect the outcome to change?
This model posits a linear relationship between covariates, treatment, and outcome; 
if the true relationship is not linear, then linear regression gives the best linear approximation to the conditional expectation of the outcome.
It is standard to use analysis of covariance (ANCOVA) to incorporate stratification in a linear model.
ANCOVA is a particular case of linear regression that allows the mean outcome to vary from stratum to stratum.
This amounts to fitting a plane to each stratum, with the constraint that they have a common slope.

Hypothesis testing of estimated coefficients requires even stronger assumptions which are not guaranteed by the experimental design.
When the assumptions hold, a hypothesis test for a treatment effect amounts to a hypothesis test of the coefficient for treatment in the ANCOVA model, and can be evaluated analytically and efficiently.
A fully saturated linear model can yield asymptotically consistent estimates and confidence intervals (\cite{lin_agnostic_2013}).
This is especially problematic in medical trials, where conditions are not always ideal for the linear model to work well:
linear models can have substantial bias in small samples (\cite{freedman_regression_2008}),
outcomes are often discrete or ordinal, 
the treatment may have a differential effect across subgroups of individuals, 
and when the distributions of outcomes may differ across strata.
%Furthermore, the hypothesis test implicitly assumes that the data were sampled at random from some underlying population, when in fact,
%medical experimenters rarely recruit patients this way (\cite{ludbrook_why_1998}).

@@

\subsection{Notation}
Suppose we have a finite population of $N$ individuals.
Individuals are grouped into strata indexed by $j = 1, \dots, J$, with $n_j$ individuals in stratum $j$ and $\sum_{j=1}^J n_j = N$.
Of the $n_j$ subjects in stratum $j$, $n_j^T$ are assigned treatment 1, while the remaining $n_j - n_j^T$ are assigned treatment 0.
Let $Z_{ij}$ indicate the treatment assigned to individual $i$ in stratum $j$.

@@

Furthermore, we observe a covariate $X_{ij}$ that may be associated with the outcome.
For expository clarity, we suppose that $X$ is univariate, but all results are easily extended to the case when $X$ is multivariate.
$X$ may be associated with stratum membership.

We are interested in the effect of treatment, measured as differences in potential outcomes $Y_{ij}(1) - Y_{ij}(0)$.
We can never learn this difference for any particular individual.
However, a tractable problem is to estimate the mean difference in the study sample or in a target population.
Various functions of potential outcomes may be of clinical interest; the goal of the study and the method of analysis determine which function is considered.

We study hypothesis testing for whether these differences are nonzero using parametric ANCOVA and its permutation counterparts, assuming this potential outcomes framework throughout.
Other valid methods for comparing two groups include using a two-sample $t$ test to test the difference in two means from normal distributions,
the Wilcoxon rank sum test to test for differences in the medians of two independent groups, 
and the Kolmogorov-Smirnov test and receiver operating characteristic curve analyses to test whether two groups have different distribution functions (\cite{lehmann_nonparametrics_1975,  vexler_statistical_2016}).
We focus on testing using the linear model as this is standard in clinical trials, 
requires fewer distributional assumptions on the data when using the potential outcomes framework, 
deals with averages,
and incorporates control variables to increase power.

@@

To carry out the standard parametric hypothesis test for a linear model, the following assumptions are needed (\cite{freedman_statistical_2005}):
\begin{itemize}
\item \textbf{Linearity:} The data $Y$ are related to $X$ and $Z$ linearly.
\item \textbf{Constant slopes:} Stratum membership only affects the intercept $\alpha_j$, not the slopes $\beta$ and $\gamma$.
\item \textbf{IID Errors:} The $\varepsilon_{ij}$ are independent and identically distributed with mean $0$ and common variance $\sigma^2$.
\item \textbf{Independence:} If $X$ is random, $\varepsilon$ is statistically independent of $X$.
\item \textbf{Normality:} The errors are normally distributed.
\end{itemize}

The coefficients are estimated using least squares (or equivalently, by maximizing the likelihood).
The coefficient $\hat{\gamma}$ is the estimated average treatment effect. 
This procedure also yields an estimate $\hat{\sigma}_{\hat{\gamma}}^2$ of the variance of $\hat{\gamma}$.
Under the null hypothesis, the test statistic 
$$ T = \frac{\hat{\gamma}}{\sqrt{ \hat{\sigma}_{\hat{\gamma}}^2}}$$
follows the Student $t$ distribution with degrees of freedom equal to the number of observations minus the number of parameters estimated (in this case, $N - J - 2$).
The $p$-value for this hypothesis test is the probability, assuming the null hypothesis of zero coefficient is true, that a value drawn from the $t$ distribution is larger in magnitude than the observed $T$.
This test is equivalent to the $F$ test.
When the model assumptions are true, it is uniformly most powerful.

The linear model is robust to violations of its assumptions, but theoretical guarantees tend to be asymptotic.
When randomized experiments are analyzed in the parametric framework, which assumes that treatment assignment is fixed and the errors are random,
the estimated treatment effect $\hat{\gamma}$ and nominal standard errors $\hat{\sigma}_{\hat{\gamma}}$ can be severely biased (\cite{freedman_regression_2008, lin_agnostic_2013}).
\cite{lin_agnostic_2013} shows that running a regression with a full set of treatment and covariate interaction terms cannot hurt asymptotic precision, and using the Huber-White sandwich standard errors can yield asymptotically valid confidence intervals.
In small samples, the bias may still be substantial.
\cite{miratrix_adjusting_2013} show that post-stratification, estimating treatment effects within groups of similar individuals defined after data are collected, can ameliorate this bias.
This is equivalent to estimating a fully saturated linear model with interaction terms for treatment and stratum membership.
The linear ANCOVA model does not account for variation in the treatment effect across strata in this way;
if the difference in potential outcomes is heterogeneous across strata, then the coefficient $\gamma$ may be attenuated towards 0.
Thus, it is unclear for any particular dataset whether or not the ANCOVA model will give valid results.

\subsection{Stratified permutation test}\label{subsec:strat_perm_test}
All permutation tests essentially have two requirements: a conditioning space, the orbit of a finite group of transformations of the data in which all configurations of the data are equiprobable under the null hypothesis, and a set of sufficient statistics for the data which describe the data under these transformations.(\cite{pesarin_permutation_2010}).
%We describe these two elements in the context of a stratified randomized experiment.
In experiments, the only random quantities are those involving $Z$, the vector of treatment assignments.
In particular, the potential outcomes and covariates are fixed in the finite population under study, while the observed responses change with $Z$.
The exact permutation inference is derived from the conditioning space which includes all possible values of $Z$.

@@

In an experiment with complete randomization, any of the possible allocations of treatment that assign $n_j^T$ out of $n_j$ individuals to treatment 1 in stratum $j$, for each stratum $j=1,\dots,J$, has equal probability.
We can construct the permutation distribution of any statistic under the null hypothesis by imputing the unobserved potential outcomes using the sharp null hypothesis and by re-randomizing treatment assignments using this principle of equal probabilities.
The experimental design specifies the permutation scheme; no additional assumptions are needed to find the distribution of any statistic.
In practice, the $\prod_{j=1}^J {n_j \choose n_j^T}$ possible allocations of treatment may be too numerous to compute the statistic for each one.
Instead, we typically use Monte Carlo simulations to randomly assign treatment and compute the statistic of interest a large number of times to approximate the randomization distribution to a desired degree of accuracy.

To summarize, the assumptions for the stratified permutation test are the following.
\begin{itemize}
\item \textbf{Complete randomization:} the $\prod_{j=1}^J {n_j \choose n_j^T}$ assignments of treatment which assign $n_j^T$ out of $n_j$ individuals to treatment 1 in stratum $j$, for each stratum $j=1,\dots,J$, has equal probability
\item \textbf{No treatment effect under the null:} both potential outcomes are identical and equal to the observed outcome for each individual $(i, j)$
\end{itemize}

@@

To avoid this, one may want to account for the stratification. 
One way to construct a stratified test is to use the same stratified permutation scheme but a different statistic which combines the stratum-specific statistics into a single value, for instance taking the sum of their absolute values.
Taking the absolute value before summing ensures that effects with different signs do not cancel each other out.
Another way to construct a test is to use the nonparametric combination (NPC) framework proposed by \citet{pesarin_permutation_2010}.
In this framework, the ``global'' null hypothesis of no effect whatsoever is decomposed into the intersection of ``partial'' null hypotheses of no treatment effect within each stratum.
For NPC, one first conducts each partial test separately, then combines their $p$-values in a way that preserves dependencies.
For a randomized experiment, this method is equivalent to combining stratum statistics directly, because treatment is independently assigned across strata.

\subsection{Permutation tests with the linear model}\label{subsec:lm_perm_tests}

The tests in Section~\ref{subsec:strat_perm_test} only use information on the treatment and the outcome.
However, experimenters typically record additional covariates that may be predictive of the outcome.
Incorporating these covariates can reduce variation in the statistic due to factors unrelated to treatment, thereby increasing power to detect a treatment effect.
The permutation tests in this section use the linear model in Equation~\ref{eqn:ancova} to control for covariates, but make different assumptions about the data.
In a randomized experiment, the treatment assignment is independent of covariates, errors, and potential outcomes,
making several variables exchangeable.
We show two permutation tests developed in this framework.
They test the same ``sharp'' null hypothesis of no treatment effect.

The first linear model permutation test is a variation on the stratified permutation test described in Section~\ref{subsec:strat_perm_test}.
The assumptions are the same.
The only difference is the test statistic:
this test uses the $t$ statistic for the regression coefficient of $Z$ in the linear model in Equation~\ref{eqn:ancova}.
Rather than using the $t$ distribution to calculate the significance of the observed $t$ statistic, we construct the permutation distribution of the $t$ statistic by simulation.
Again, the experimental design specifies the randomization distribution and no additional assumptions are needed:
$Z$ is independent of $(Y(1), Y(0), X)$ conditional on strata.

@@

If the null hypothesis is true, then $\gamma = 0$ and $\varepsilon_{ij} = Y_{ij} - \alpha_{ij} - \beta X_{ij}$ for all $i = 1, \dots, n_j, j = 1, \dots, J$.
Therefore, we may estimate the errors by $\hat{\varepsilon} =Y - \hat{Y}$, where $\hat{Y}$ is the vector of fitted values from the regression of $Y$ on $X$ but not $Z$.
The $\hat{\varepsilon}$ approximate the true errors $\varepsilon$ from the true data-generating process, assuming that the null hypothesis is true.
Furthermore, under the null hypothesis, the $\varepsilon$ are independent of $Z$ within strata.

We can carry out an approximate permutation test by permuting the estimated $\hat{\varepsilon}$ within strata, independently across strata.
The test is only approximate because we must estimate $\alpha$ and $\beta$ to construct $\hat{\varepsilon}$.
The test has two important assumptions pertaining specifically to the linear regression, on top of the two necessary pieces of a permutation test:
\begin{itemize}
\item \textbf{Linear model:} the linear model that describes $Y$ using $X$ and the approximation error $\varepsilon$ holds
\item \textbf{Linear regression fit:} the linear regression residuals $\hat{\varepsilon}$ are a good approximation to the true errors $\varepsilon$ 
\item \textbf{Conditioning space:} the $\varepsilon$ are exchangeable against $X$ within strata
\item \textbf{No treatment effect under the null:} the $\varepsilon$ do not depend on $Z$
\end{itemize}

The $\hat{\varepsilon}$ may not even be approximately exchangeable within strata if the regression does not describe the relation between $Y$ and $X$ well.
For this reason, \cite{freedman_nonstochastic_1983} advise against using the method if there are large outlier values in the covariate $X$ or when $X$ and $Z$ are highly colinear.
Randomization does not guarantee the exchangeability against $X$, so this assumption should be checked.
One way, suggested by \cite{freedman_nonstochastic_1983}, is to visually inspect a scatterplot of the residuals against $X$.
One may also test for association between $X$ and $\hat{\varepsilon}$, such as a stratified permutation test using the correlation as a statistic, independently within each stratum and use the nonparametric combination of tests to obtain a single $p$-value (\cite{pesarin_permutation_2010}).

@@

We simulated data from a randomized experiment using several different data-generating processes.
The variables included stratification, two randomly assigned treatments, a baseline measure before treatment, and the same variable measured after treatment.
We tested for a treatment effect and compared the empirical power of the tests over repeated random treatment assignments.
We compared the following five tests:
the $t$ test from the parametric ANCOVA,
a stratified permutation test using the difference in mean outcomes
 (called ``stratified permutation'' in what follows),
a stratified permutation test using the mean change scores, defined as the difference between the outcome and baseline measure (called ``differenced permutation'' in what follows),
a stratified permutation test based on the $t$ statistic from the linear regression of outcome on control variables (called ``LM permutation'' in what follows),
and the Freedman-Lane permutation test.
The simulation code is included in four supplementary files.

@@

In the first set of simulations, we generated continuous potential outcomes based on a latent variable and disturbance terms, both of which we manipulated.
We used two scenarios.
In the first scenario, the treatment effect was constant across strata.
We drew a latent random variate $v_{ij}$ from the uniform distribution on $[-4, 4]$.
In the second scenario, the treatment effect varied across strata, and we drew the latent random variable according to
\begin{displaymath}
   v_{ij} \sim \left\{
     \begin{array}{lr}
       \text{Unif}[-4, -1] & : j=1\\
       \text{Unif}[-1, 1] & : j=2\\
       \text{Unif}[1, 4] & : j=3\\
     \end{array}
   \right.
\end{displaymath}

@@

We included $n_j=16$ individuals per stratum and treatment assignment was balanced, i.e. 8 people received each treatment in each stratum.
After sampling $(v_{ij}, \varepsilon_{ij}, \delta_{ij})$, we constructed the baseline value for individual $i,j$ as 
$$ X_{ij} = \frac{-\gamma e^{v_{ij}} + e^{v_{ij}/2}}{2} + \varepsilon_{ij}.$$

\noindent $\gamma$ was a fixed constant.
Then we generated potential outcomes as

@@

\noindent This is equivalent to the following linear data-generating process:

$$Y_{ij}(Z_{ij}) = Z_{ij} \gamma e^{v_{ij}} + X_{ij} + (\delta_{ij} - \varepsilon_{ij}).$$

\noindent This model does not exactly match the ANCOVA Equation~\ref{eqn:ancova} and only some of the assumptions are satisfied:
$Y$ is linear in $(X, Z)$ conditionally on the latent variable $v$, 
the errors are independent and identically distributed, and are independent of $X$.
When the latent variables vary in distribution across strata, the assumption of constant slopes is violated.

We regenerated $Z$ and the potential outcomes $Y(Z)$ 10,000 times.
We repeated this procedure for each distribution of latent variables $v$ and of the errors $\varepsilon$ and $\delta$.
The treatment effect for individual $(i, j)$ was $\gamma e^{v_{ij}}$.
We tested whether the treatment had a nonzero effect in the sample using each of the tests described.

\begin{figure}[h]
\centering
\includegraphics[width = \textwidth]{fig/continuous_outcomes_simulation_power.pdf}
\caption{Empirical power curves for the continuous simulated data with normally distributed $\varepsilon$ and $\delta$. The left panel shows power when the treatment effect was constant across strata and the right panel shows power when the treatment effect was heterogeneous.}
\label{fig:continuous_outcomes_sim_power}
\end{figure}

First, we studied how power changed with the average treatment effect.
We used normally distributed errors and varied $\gamma$ from $0$ to $0.2$ in steps of $0.05$.
In the random populations that were generated, this corresponded to population average treatment effects of $0, 0.34, 0.68, 1.02$, and $1.36$ respectively.
The left panel of Figure~\ref{fig:continuous_outcomes_sim_power} shows the empirical power (rate of rejection in the 10,000 simulations) at level 5\% for these increasing effect sizes.
When the treatment effect was homogeneous across strata, all five tests had the correct level of 5\% (i.e. the power when $\gamma = 0$).
Power of the differenced permutation test increased at a slower rate than the other tests.
This is unsurprising, as the correlation between baseline $X$ and outcome $Y$ was low:
its highest value was $0.33$ when $\gamma=0$ and it decreased to nearly $0$ when $\gamma=0.2$.
The parametric and permutation linear model tests performed comparably and are plotted on top of each other, while the stratified permutation test had slightly more power.
